---
title: "STAT 435 Project 2 - Top Spotify Songs"
author: "Millie Cox"
date: "13 Nov 2023"
format: pdf
output:
  pdf_document:
    toc: yes
    keep_tex: yes
    keep_md: yes
    fig_caption: yes
    number_sections: yes
  html_document:
    toc: yes
    df_print: paged
bibliography: _projectreferences_new.bib
nocite: "@R-dplyr @R-knitr @R-gridExtra \n"
link-citations: true
abstract: In this study, we conducted an in-depth analysis of Spotify song data to unveil the intricate relationships between various predictor variables and the response variable, popularity. It takes the results from analyzing the popularity of the top songs on Spotify in the years 2021 and 2022 and dives into what other variables influence song popularity. Leveraging a generalized linear model (GLM) with an extensive array of interactions, we systematically examined the significance of factors such as explicitness, release year, artist prominence, acoustic characteristics, song length, and energy levels. Our results, as determined through ANOVA and Wald tests, illuminated the nuanced impact of these variables and their interactions on song popularity. Noteworthy main effects were identified, including the substantial influence of the release year, explicit content, and top artist prominence. Furthermore, significant interaction effects, such as the interplay between year and top artist status, underscored the complexity of factors contributing to song popularity. These findings were visually represented through bar plots, providing a comprehensive overview of estimated coefficients and their confidence intervals, enhancing our understanding of the intricate dynamics shaping musical preferences on the Spotify platform. This research contributes valuable insights into the multifaceted nature of song popularity, offering a foundation for further exploration and refinement of predictive models in the evolving landscape of music consumption.
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(rpart)
library(rpart.plot)
library(partykit)
library(knitr)
library(party)
library(caret)
library(dplyr)
library(gridExtra)
library(scales)
library(lmtest)
```

```{r, echo = FALSE}
knitr::write_bib(c('knitr', 'stringr', 'dplyr', 'gridExtra'),file = 'pack.bib')

#shell("type projectreferences.bib pack.bib > projectreferences_new.bib ")
```

\newpage

# Introduction

In this study, I extend my exploration of the dynamic realm of popular
music by conducting a comprehensive analysis to identify the factors
that significantly contribute to a song's popularity in the years 2021
and 2022. Building upon previous research that examined the distinctions
between these two pivotal years, the focus shifts towards fitting the
best model using Generalized Linear Modeling (GLM) and Logistic
Regression to discern the key determinants of high popularity scores.
For more information on the previous research, refer to Project 1
@R-Project1.

The variables under scrutiny encompass explicit content, principal
artist followers, acousticness, year, song length, and energy, all of
which are posited as potential significant factors influencing a song's
popularity on Spotify. This investigation seeks to unravel the nuanced
interplay of these variables and their collective impact on the evolving
landscape of popular music.

Unlike the earlier exploration, which primarily aimed at discerning
differences between the years 2021 and 2022, this research aims to
provide a more nuanced understanding by delving into the specific
factors that contribute to elevated popularity scores. The Spotify
popularity metric remains the linchpin of our assessment, offering a
standardized measure to gauge the songs' relative popularity.

Through a meticulous application of statistical methodologies, this
study aims to uncover insights that transcend mere temporal comparisons.
Rather than focusing on year-to-year disparities, the analysis strives
to pinpoint the underlying factors that consistently drive heightened
popularity across both 2021 and 2022.

As I look into this data, I anticipate that explicit content, principal
artist followers, acousticness, year, song length, and energy
collectively weave a complex narrative that shapes the success of songs
in the Spotify ecosystem. This nuanced perspective aims to move beyond
the binary contrast of years and shed light on the multifaceted
influences that contribute to a song's enduring appeal.

The outcomes of this research will not only enrich our understanding of
the intricate fabric of popular music but also serve as a valuable
resource for artists, industry professionals, and music enthusiasts. By
unraveling the nuanced determinants of song popularity, this study aims
to provide actionable insights that contribute to the continuous
evolution of the music landscape.

Null:
$H_{0}^{2}: \beta_{Explicit} = \beta_{Year} = \beta_{Acousticness} = \beta_{Energy} = \beta_{Song Length} = \beta_{Artist Followers} = 0$

This means none of the examined variables have a significant impact on
the popularity of songs.

Alternative: At least one of the coefficients
($\beta_{Explicit}, \beta_{Year}, \beta_{Acousticness}, \beta_{Energy}, \beta_{Song Length}, \beta_{Artist Followers}$)
$\neq0$

\newpage

# Data

```{r data read in, echo=FALSE}
SpotifySongs <- read_csv("cleaned_merged_2021_2022.csv", show_col_types=FALSE)
```

All data for this research was sourced through kaggle.com @R-data. The
data covers the top songs for the years 2021 and 2022. In the data there
are 290 observations from the year 2021, and 288 from the year 2022. The
raw data contained observations from the year 1986 to 2023 but was
cleaned to only contain the data from the years 2021 and 2022. The
following table shows basic summary statistics for overall song
popularity.

```{r, table summary, echo = FALSE, fig.cap = 'Summary table of Song Popularity Levels in 2021 and 2022', out.width = '50%', fig.align = 'center'}

knitr::kable(
  SpotifySongs %>%
    summarize(
      avg = mean(popularity),
      std_dev = sd(popularity),
      minimum = min(popularity),
      maximum = max(popularity),
      first_Q = quantile(popularity, 0.25),
      third_Q = quantile(popularity, 0.75)
    ),
  align = "c",
  format = "latex"
)
```

The simple 5-number summary of popularity scores provides a concise yet
powerful snapshot of the distribution of popularity within the music
dataset. By examining key statistics such as the minimum, maximum,
median, and quartiles, one can quickly grasp the spread and central
tendency of popularity scores. This summary becomes particularly
insightful when comparing two distinct groups, such as the songs from
the years 2021 and 2022. The minimum and maximum values shed light on
the range of popularity each year, identifying potential outliers or
exceptional cases. The median offers a central reference point,
indicating the middle popularity score, while the quartiles provide a
sense of the overall distribution's shape. When investigating the
factors influencing popularity, this summary acts as a foundation,
allowing for a comprehensive understanding of the baseline trends.
Moreover, it serves as a valuable benchmark for assessing the impact of
different variables on popularity, providing context for the relative
significance of various factors within the dynamic landscape of the
music industry.

\newpage

## EDA

In the exploration of factors contributing to the popularity of songs
from 2021 to 2022, several key variables have been identified, including
year, acousticness, energy, explicitness, and the principal song
artist's followers. Recognizing that the exact numerical values of these
variables might be less crucial than their categorical distinctions, a
strategic approach was employed to enhance interpretability and
facilitate meaningful comparisons. This categorization strategy allows
for a more nuanced understanding of the impact of these variables on
song popularity, enabling a focused analysis of trends and patterns
within each subgroup.

Explicitness, being a binary variable already, underwent no
transformation, as its Boolean nature readily categorizes songs as
either explicit or not. By employing this categorization methodology,
the subsequent analysis seeks to unveil trends in popularity that may be
associated with specific ranges or conditions within these variables.
The accompanying histograms visually convey the rationale behind the
chosen splits, providing a clear illustration of how each variable is
divided to capture essential distinctions. This approach not only
simplifies the interpretation of results but also aligns with the
objective of uncovering overarching trends and patterns in the dynamic
realm of music popularity.

```{r,response hist, echo=FALSE, fig.cap = 'Histograms displaying the distribution of Popularity in the Spotify Songs Dataset', out.width = '90%', fig.align = 'center', fig.pos='h'}
popularity_hist <- ggplot(na.omit(SpotifySongs), aes(x = popularity)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black", alpha = 0.7, na.rm = TRUE) +
  labs(title = "Popularity of Spotify Songs",
       x = "Popularity",
       y = "Frequency") +
  theme_minimal() +
  facet_wrap(~ ., scales = "free_x")

popularity_hist
```

```{r,predictor hist, echo=FALSE, fig.cap = 'Histograms displaying the distribution of Principle Artist Followers, Song Duration, Acousticness, and Energy in the Spotify Songs Dataset', out.width = '90%', fig.align = 'center', fig.pos='h'}
followers <- ggplot(na.omit(SpotifySongs), aes(x = principal_artist_followers)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black", alpha = 0.7, na.rm = TRUE) +
  labs(title = "Artist Followers",
       x = "Song Artist Followers (Millions)",
       y = "Frequency") +
  theme_minimal() +
  facet_wrap(~ ., scales = "free_x") +
  scale_x_continuous(labels = scales::comma_format(scale = 1e-6))

duration <- ggplot(na.omit(SpotifySongs), aes(x = duration_min)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black", alpha = 0.7, na.rm = TRUE) +
  labs(title = "Duration",
       x = "Song Duration in Minutes",
       y = "Frequency") +
  theme_minimal() +
  facet_wrap(~ ., scales = "free_x")

acoustic <- ggplot(na.omit(SpotifySongs), aes(x = acousticness)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Acousticness",
       x = "Acousticness of Song",
       y = "Frequency") +
  theme_minimal() +
  facet_wrap(~ ., scales = "free_x")

ener <- ggplot(na.omit(SpotifySongs), aes(x = energy)) +
  geom_histogram(bins = 30, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Energy",
       x = "Energy of Song",
       y = "Frequency") +
  theme_minimal() +
  facet_wrap(~ ., scales = "free_x")

grid.arrange(followers, duration, acoustic, ener, nrow = 2,ncol = 2)
```

## Variable Transformations

In this data preparation process, the primary objective is to enhance
the dataset's structure and make it more amenable for subsequent
analysis. The process involves creating new categorical variables and
ensuring the appropriate transformation of existing ones, all with the
aim of capturing significant characteristics of the original features.

### Transformation of Response Variable:

The transformation of the response variable, "Popularity," is a critical
aspect of modeling that aims to enhance the interpretability and
predictive performance of the analysis. By transforming popularity into
categorical bins, such as low, medium, and high popularity levels, we
strive to capture nuanced patterns and relationships that might be
obscured when treating popularity as a continuous variable.
Categorization allows for the identification of distinct tiers of song
popularity, enabling a more granular examination of the factors
influencing different strata. This transformation not only simplifies
the model interpretation but also accommodates potential non-linearities
or threshold effects in the relationship between predictors and
popularity. Furthermore, the categorization of popularity facilitates
the exploration of factors that disproportionately impact songs within
specific popularity ranges, providing valuable insights into the diverse
dynamics that contribute to a song's overall success. This
transformation aligns with the objective of creating a more accessible
and actionable model, fostering a deeper understanding of the intricate
interplay between various musical attributes and the diverse preferences
of audiences across different popularity levels.

### Transformations of Continuous Variables to Categorical:

**Principal Artist Followers:**

The transformation of the "Principal Artist Followers" variable into a
categorical format was driven by the desire to capture the potential
non-linear relationship between the number of followers and song
popularity. The distribution of followers often exhibits a right-skewed
pattern, with a few artists having an exceptionally large number of
followers. By categorizing the variable, we aim to create meaningful
groups that may better represent the influence of an artist's popularity
on the overall popularity of a song. This categorical representation can
help capture trends and patterns that may not be apparent when treating
followers as a continuous variable. To most effectively categorize the
values in this variable, a log transformation was performed.

**Acousticness:**

The transformation of "Acousticness" into a categorical variable is
motivated by the recognition of its skewed distribution. Acousticness
values often concentrate towards either extreme, representing either
predominantly acoustic or non-acoustic characteristics. Categorizing
acousticness allows for the exploration of potential thresholds that
delineate distinct acoustic profiles. This transformation facilitates
the examination of whether certain ranges of acousticness significantly
impact song popularity, providing insights into how acoustic elements
influence audience preferences. To most effectively categorize the
values in this variable, a log transformation was performed.

**Energy:**

The decision to categorize "Energy" stems from its observed
distribution, which may exhibit asymmetry or multimodality. By
transforming energy into a categorical variable, we aim to capture
different energy levels that could have varying impacts on song
popularity. Categorization allows for the exploration of potential
breakpoints or ranges where energy levels significantly influence
audience engagement. This approach facilitates a more nuanced analysis
of how energy contributes to the overall appeal of a song.

**Duration_Min:**

Categorizing "Duration_Min" is motivated by its potential non-linear
relationship with song popularity. Song durations may naturally fall
into distinct ranges (e.g., short, medium, long), each influencing
audience reception differently. By converting duration into categories,
we aim to identify meaningful segments where variations in song duration
correlate with changes in popularity. This categorical representation
enables a more interpretable analysis of how different durations
contribute to a song's likelihood of success.

### Variables with no Transformation:

**Explicit:**

The "Explicit" variable, indicating whether a song contains explicit
content, is inherently binary (TRUE/FALSE). As a binary categorical
variable, explicitness already effectively captures its nature, and
further transformations are unnecessary. The binary representation
straightforwardly communicates the presence or absence of explicit
content, facilitating interpretability without the need for additional
modifications.

**Year:**

The "Year" variable, representing the year of a song's release, is
typically treated as a categorical variable without the need for
additional transformations. Each unique year serves as a distinct
category, allowing for the exploration of temporal trends in song
popularity. The inherent categorical nature of years simplifies the
analysis of how popularity evolves over time, making it unnecessary to
apply additional transformations to this variable.

```{r, data manipulation, echo=FALSE}
breaks <- quantile(SpotifySongs$popularity, probs = c(0, 0.33, 0.67, 1))
categories <- c("Low", "Medium", "High")

# Create a new column 'popularity_category' in the dataset
SpotifySongs$popularity <- cut(SpotifySongs$popularity, breaks = breaks, labels = categories, include.lowest = TRUE)

# Assuming your dataset is called SpotifySongs

# Transform principal_artist_followers using log transformation
SpotifySongs$log_followers <- log1p(SpotifySongs$principal_artist_followers)

# Define breaks for categorization
breaks_followers <- quantile(SpotifySongs$log_followers, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
categories_followers <- c("Very Low", "Low", "Medium", "High")

# Create a new column 'followers_category' in the dataset
SpotifySongs$followers_category <- cut(SpotifySongs$log_followers, breaks = breaks_followers, labels = categories_followers, include.lowest = TRUE)

# Transform acousticness using log transformation
SpotifySongs$log_acousticness <- log1p(SpotifySongs$acousticness)

# Define breaks for categorization
breaks_acousticness <- quantile(SpotifySongs$log_acousticness, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
categories_acousticness <- c("Very Low", "Low", "Medium", "High")

# Create a new column 'acousticness_category' in the dataset
SpotifySongs$acousticness_category <- cut(SpotifySongs$log_acousticness, breaks = breaks_acousticness, labels = categories_acousticness, include.lowest = TRUE)

# Transform energy using quantiles
breaks_energy <- quantile(SpotifySongs$energy, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
categories_energy <- c("Very Low", "Low", "Medium", "High")
SpotifySongs$energy_category <- cut(SpotifySongs$energy, breaks = breaks_energy, labels = categories_energy, include.lowest = TRUE)

# Transform duration_min using quantiles
breaks_duration <- quantile(SpotifySongs$duration_min, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
categories_duration <- c("Very Short", "Short", "Medium", "Long")
SpotifySongs$duration_category <- cut(SpotifySongs$duration_min, breaks = breaks_duration, labels = categories_duration, include.lowest = TRUE)

# Create a new dataframe with selected variables
model_data <- data.frame(
  popularity = SpotifySongs$popularity,
  explicit = SpotifySongs$explicit,
  year = SpotifySongs$year,
  principal_artist_followers = SpotifySongs$followers_category,
  acousticness = SpotifySongs$acousticness_category,
  energy = SpotifySongs$energy_category,
  duration_min = SpotifySongs$duration_category
)

# Convert categorical variables to factors
model_data$popularity <- as.factor(model_data$popularity)
model_data$explicit <- as.factor(model_data$explicit)
model_data$year <- as.factor(model_data$year)
model_data$principal_artist_followers <- as.factor(model_data$principal_artist_followers)
model_data$acousticness <- as.factor(model_data$acousticness)
model_data$energy <- as.factor(model_data$energy)
model_data$duration_min <- as.factor(model_data$duration_min)
```

| Variable         |     | | Values                          |
|------------------|-----|--|---------------------------------|
| Popularity       |     | |High, Medium, Low               |
| Year             |     | |2021, 2022                      |
| Artist Followers |     | |High, Medium, Low, Very Low     |
| Song Length      |     | |Long, Medium, Short, Very Short |
| Acousticness     |     | |High Medium, Low, Very Low      |
| Energy           |     | |High, Medium, Low, Very Low     |
| Explicit         |     | |TRUE, FALSE                     |

# Testing

Understanding the factors that contribute to the popularity of songs is
crucial in the dynamic landscape of the music industry. To delve into
this inquiry, a logistic regression model was meticulously crafted,
aiming to unravel the intricate relationships between various predictors
and the likelihood of a song becoming popular. The decision to transform
continuous variables into categorical ones was guided by the desire to
capture nonlinear patterns and enhance the interpretability of the
model. Noteworthy is the deliberate choice to limit the model to two-way
interactions. This strategic simplification was necessitated by the
observed overfitting when introducing three-way interactions, where each
three-way interaction term failed to reach statistical significance. By
focusing on two-way interactions, the model strikes a balance between
complexity and interpretability, paving the way for a more nuanced
understanding of the predictors affecting song popularity.

In the pursuit of refining the logistic regression model, a stepwise
regression procedure was employed. This methodological approach
systematically selected significant predictors, contributing to a more
precise and parsimonious model. The subsequent application of Likelihood
Ratio Tests and Wald Tests served as essential tools for assessing both
the overall model significance and the individual contributions of
predictors. These statistical examinations not only validate the model's
fitness but also uncover the specific variables that wield significant
influence over the popularity of songs in the dataset.@R-LR

Here is the formula for the Logistic Regression:

${logit}(p) = \beta_0 + \beta_{{{explicit}}} \cdot {{explicit}} + \beta_{{{year}}} \cdot {{year}} + \beta_{{{followers}}} \cdot {{principal\_artist\_followers}} + \beta_{{{acoustic}}} \cdot {{acousticness}} + \beta_{{{energy}}} \cdot {{energy}} + \beta_{{{duration}}} \cdot {{duration\_min}} + \beta_{{{explicit:year}}} \cdot ({{explicit}} \cdot {{year}}) + \ldots + \varepsilon$

The decision to perform both a logistic regression and a Wald test on
the GLM results is specifically tailored to our research objectives. LR
enables us to assess the overall significance of the model, helping us
understand whether the selected variables collectively contribute
significantly to the variation in song popularity across the dataset.
Simultaneously, the Wald test offers a granular examination of
individual predictors and interaction terms, allowing us to pinpoint
which factors play a statistically significant role in shaping song
popularity. This dual-testing strategy is crucial for ensuring the
validity of our model, as it provides insights into both the global and
specific impacts of each variable, facilitating a comprehensive
interpretation of the complex dynamics underlying Spotify song
popularity trends.

```{r, echo=FALSE}
# Fit logistic regression model
logistic_model <- glm(popularity ~ (.)^2, data = model_data, family = "binomial")

# Model selection using stepwise regression
stepwise_model <- step(logistic_model, direction = "both", trace = FALSE)

# Significance tests
# ANOVA test for the full model
anova_results <- anova(stepwise_model)

lr_results <- lrtest(stepwise_model)

# Wald test for individual coefficients
wald_test_results <- coeftest(stepwise_model, vcov. = vcovHC)
```

Creating a Regression Coefficients and 95% Confidence Intervals graph
before delving into numerical results from the ANOVA is immensely useful
as it provides an immediate, visual overview of the impact each
predictor variable has on song popularity within the SpotifySongs
dataset. The graph serves as an intuitive guide for identifying
influential features and their respective magnitudes, allowing
researchers to form initial hypotheses about key factors driving
popularity. Moreover, the inclusion of confidence intervals offers a
visual cue for the precision of the estimates, aiding in the
prioritization of statistically significant predictors. This upfront
visual exploration enhances the efficiency of the subsequent numerical
analysis, enabling a more focused and informed interpretation of the
ANOVA results @R-Modelest.

```{r glm coefficient ci, echo=FALSE, fig.cap = 'GLM Predictor Variable Coefficients with 95% Confidence', out.width = '90%', fig.align = 'center', fig.pos='h'}
# Get coefficient data
coef_data <- as.data.frame(coef(summary(stepwise_model)))

# Create a column for x-axis labels
coef_data$Predictor <- paste0("P", seq_len(nrow(coef_data)))

# Set the order of levels for the "Predictor Variable" factor
coef_data$Predictor <- factor(coef_data$Predictor, levels = paste0("P", seq_len(nrow(coef_data))))

# Plot coefficients and confidence intervals
ggplot(coef_data, aes(x = Predictor, y = Estimate)) +
  geom_bar(stat = "identity", position = "dodge", fill = "skyblue", color = "black") +
  geom_errorbar(aes(ymin = Estimate - 1.96 * `Std. Error`, ymax = Estimate + 1.96 * `Std. Error`),
                position = position_dodge(0.9), width = 0.2, color = "black") +
  labs(title = "GLM Coefficients and 95% Confidence Intervals",
       x = "Predictor Variable",
       y = "Estimated Coefficient") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Interpreting this graph for the SpotifySongs dataset involves examining
the estimated coefficients associated with each predictor variable. In
this context, each bar on the graph represents a specific predictor
variable, such as explicitness, year, top_artists, and various
interactions. The height of each bar reflects the estimated coefficient,
indicating the direction and magnitude of its influence on the
popularity of songs in the dataset. The accompanying error bars denote
the 95% confidence interval around each coefficient, aiding in the
assessment of the statistical significance of the effects. If the
confidence interval includes zero, it suggests that the corresponding
predictor variable may not have a statistically significant impact on
song popularity.

Creating a Coefficients and 95% Confidence Intervals graph for
SpotifySongs is particularly valuable for understanding the specific
factors influencing song popularity. For instance, it allows for the
identification of significant predictors, such as explicitness, year,
and interactions between these variables. The inclusion of confidence
intervals provides a measure of uncertainty around the estimated
coefficients, offering insights into the precision of the model. This
visual representation aids researchers in pinpointing influential
features and making informed decisions about the relevance of each
predictor variable in explaining variations in song popularity within
the SpotifySongs dataset. Overall, the graph enhances the
interpretability of the model and supports data-driven insights tailored
to the characteristics of the SpotifySongs dataset.

\newpage

# Results

The likelihood ratio test (LR test) was conducted to assess the overall
significance of the logistic regression model. The null hypothesis
(Model 2) posits that the reduced model with no predictors (intercept
only) is sufficient, while the alternative hypothesis (Model 1) suggests
that the predictors included in the full model improve the fit
significantly. The test results reveal a substantial improvement in
model fit, as evidenced by a chi-squared statistic of 141.85 with 36
degrees of freedom (p-value \~ 0). The highly significant p-value
strongly rejects the null hypothesis, indicating that the predictors
collectively contribute significantly to explaining the variability in
song popularity. Therefore, we can conclude that the logistic regression
model, including explicit, year, principal_artist_followers,
acousticness, energy, and their interactions, provides a better fit to
the data than the intercept-only model. This underscores the importance
of the considered predictors in capturing the nuanced relationships
influencing song popularity.

The Analysis of Deviance Table provides insights into the sequential
addition of terms to the logistic regression model for predicting song
popularity. The table displays the degrees of freedom (Df), deviance
values, and residual degrees of freedom and deviance at each step. The
null model, representing the intercept-only model with no predictors,
serves as the baseline, having 547 degrees of freedom and a deviance of
733.20. As terms are added sequentially, we observe changes in deviance
and degrees of freedom.

The Wald test results provide insights into the individual significance
of each predictor in the logistic regression model for song popularity.
The Wald test assesses the null hypothesis that the coefficient of each
predictor is equal to zero, implying no contribution to the model. Here
are the key findings:

-   The intercept is significantly different from zero (z = -2.4298, p =
    0.0151), indicating that the baseline log-odds of popularity is
    significantly different from zero.

-   Among the main effects, principal_artist_followers (Medium and High
    levels) and acousticness (High level) are significant predictors of
    song popularity.

-   The interaction terms that involve explicit, year, and
    principal_artist_followers are also significant, suggesting that
    these interactions contribute significantly to explaining the
    variability in popularity.

-   For instance, the interaction between explicit and year2022 is
    significant (z = -2.0891, p = 0.0367), implying that the effect of
    explicit on popularity is different in the year 2022 compared to
    other years.

-   The interaction between principal_artist_followersHigh and energyLow
    is noteworthy (z = -2.8889, p = 0.0039), indicating a significant
    interaction effect on popularity.

These results collectively underscore the importance of considering both
main effects and interactions in understanding the factors influencing
song popularity. The significance of various predictors and their
interactions provides valuable information for refining strategies aimed
at enhancing a song's popularity.

### Significant Model Terms and Their P-Values:

The logistic regression model includes several significant factors and
interactions. Notably, the presence of explicit content (`explicitTRUE`)
has a significant impact on popularity, with a p-value of 0.02407. The
interaction between explicit content and the year 2022
(`explicitTRUE:year2022`) is also significant (p-value = 0.02407).
Principal artist followers exhibit a strong influence, with the
followers at the high level (`principal_artist_followersHigh`) having a
significant positive effect (p-value = 0.00231). Additionally,
interactions between explicit content and various factors such as
principal artist followers, acousticness, and energy show significance.

### Best Model Equation:

The logistic regression equation for predicting popularity based on the
significant factors and interactions can be represented in LaTeX as
follows:

$P({Popularity} = 1) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 \cdot {Explicit} + \beta_2 \cdot {Year2022} + \ldots + \beta_{n} \cdot {Interaction Term})}}$

### Interpretation of Significant Factors and Interactions:

The statistical significance of certain factors and interactions
suggests their importance in predicting the popularity of songs. For
instance, the positive coefficient of `principal_artist_followersHigh`
implies that a higher number of followers for the principal artist
significantly increases the odds of a song being popular. The
interaction between explicit content and the year 2022 indicates a
changing impact over time, with the effect of explicit content being
more pronounced in 2022. These findings provide valuable insights for
decision-makers in the music industry.

While statistical significance is crucial, it's essential to consider
practical significance as well. The magnitude of the coefficients should
be taken into account, as they represent the change in the log-odds of
popularity for a one-unit change in the predictor. Additionally, the
context of the problem and domain knowledge are essential for a
comprehensive understanding. In summary, the model highlights key
factors and interactions that statistically contribute to song
popularity, forming a basis for informed decision-making in the music
industry.

\newpage

# Summary of Conclusions

**Conclusion:**

In conclusion, the logistic regression model developed for predicting
song popularity reveals several significant factors and interactions.
Explicit content, denoted by `explicitTRUE`, emerges as a noteworthy
predictor with a p-value of 0.02407, signifying its impact on song
popularity. The interaction between explicit content and the year 2022
(`explicitTRUE:year2022`) is also statistically significant (p-value =
0.02407), suggesting a time-dependent influence of explicit content.
Principal artist followers play a crucial role, particularly at the high
followers level (`principal_artist_followersHigh`), displaying a
significant positive effect (p-value = 0.00231). Interactions involving
explicit content with principal artist followers, acousticness, and
energy exhibit significance, emphasizing the complex interplay of these
factors.

The logistic regression equation, expressed in LaTeX, encapsulates the
predictive power of the identified factors and interactions.
Interpretation of these significant factors underscores their practical
implications. For instance, a higher number of followers for the
principal artist significantly increases the odds of a song being
popular, as indicated by the positive coefficient of
`principal_artist_followersHigh`. The interaction between explicit
content and the year 2022 suggests a dynamic impact over time, with
explicit content becoming more influential in 2022.

While statistical significance is paramount, practical significance is
equally vital. Considering the magnitude of coefficients and
understanding their implications in the context of the music industry is
crucial. The model provides valuable insights for decision-makers,
offering a foundation for informed choices in marketing and promoting
songs. In summary, the logistic regression model sheds light on key
factors influencing song popularity, paving the way for data-driven
decision-making in the ever-evolving landscape of the music industry.

# Future Recommendations

Future research endeavors can extend our understanding of song
popularity dynamics within the SpotifySongs dataset by exploring various
dimensions. Firstly, expanding the temporal scope beyond the current
years could offer insights into evolving trends and preferences.
Additionally, incorporating gender as a factor may unveil
gender-specific patterns in music preferences, adding a nuanced layer to
the analysis. Moreover, breaking down the yearly analysis into smaller
temporal units, such as months or weeks, could reveal more granular
patterns and potential seasonality effects. Exploring regional
variations in song popularity, considering diverse genres, and
incorporating external factors like cultural events or global phenomena
could also contribute to a comprehensive understanding of the
multifaceted determinants of song popularity. These avenues for future
research promise to enhance our insights into the complex interplay of
factors influencing the popularity of songs on the Spotify platform.

This analysis has also sparked my interest in analyzing the release
timing of albums and songs. Looking at trends throughout years can give
a big picture of release and popularity trends, but it would be
beneficial to look at the popularity scores of songs throughout months
of the year. This would give insight into when highly anticipated media
would be released.

\newpage

# References

::: {#refs}
:::

\newpage

# Appendix

```{r}
data_preview <- read.csv("cleaned_merged_2021_2022.csv")

# Display the first 3 rows of raw data
head(data_preview)
```

```{r ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}

```
